The role of lithic bipolar technology in Western Iberia’s Upper Paleolithic: the case of Vale Boi (southern Portugal)

### 2018-08-01 analysis/paper/paper.Rmd
Discussion revised

Bipolar technology clearly had an important role on the adaptive systems of the first modern humans in Western Iberia and elsewhere. As mentioned by Shot and Tostevin [-@shott_diversity_2015], a rise in bipolar technology frequency may be related to human mobility patterns and even evolutionary trends [@duke_weaving_2015]. For the first modern humans, as well as their descendants, this technology would have played a major role in the intensive and diverse exploitation of resources. In Vale Boi, bipolar technology shows up in the archaeological record connected to evidence related to the intensification and diversification of resource exploration such as the above-mentioned grease obtention techniques. These have been shown by the constant presence of impact fractures in ungulate bones, low percentage of bone areas related to higher amounts of grease, and the presence of red deer bone fragments from bones with considerably higher marrow and grease contents. Other  indicators are both the constant presence of thermally altered quartz fragments that might be linked to stone boiling activities, and a large number of greywacke slabs with impact marks revealing their use as anvils.

Our results show that in Vale Boi, scaled pieces have several patterns that stay consistent across all occupations. One relevant trend is the fact that while quartz pieces were only explored in a single axis, chert pieces were, sometimes, explored in multiple axes. One possible, rather improbable, reason for these patterns might be the fact that quartz would be more easily available than chert. Another, more likely, possibility is due to the material properties of each raw material. While with quartz it may be possible to continuously use a single axis in a piece, reducing it in only one axis without losing its efficiency, with chert it may be necessary to rotate it in order to continue obtaining usable edges. On the other hand, the higher number of axis could simply suggest that chert was differently managed and economized. Nevertheless, the simple fact that most of the pieces with multiple axes are on average larger than the rest suggests that this hypothesis is likely incorrect. From an economic standpoint, the ideal would be to continuously use the same piece without any risk of loss of the features that made it ideal for this kind of use. Throughout the use of these pieces, they suffer successive violent blows, even the platform that rests on a surface suffers some sort of heavy impact. In this specific case, we argue that there is little to no control on how the piece’s features change over its lifetime use.

From a paleoanthropological point of view, this type of bipolar technique would have a major role on a day to day basis. For carcass processing and bone marrow extraction the use of scaled pieces would bring two major advantages: (1) it would allow for a better control of bone fracture; and, in consequence, (2) it would allow a clear choice of which part of the bone can either be broken or removed. It can be argued that both activities can be performed with any kind of hammer. However, by using a hard hammer directly on the bone it may either over fracture the bone or even crush it due to the direct propagation of force (and partially damage the marrow). The use of a wedge for these activities provides more control, avoiding complete crushing of the bone. It also is much easier to cleanly extract bone marrow from a cracked bone than from a smashed bone. According to both Leblanc [-@leblanc_wedges_1992] and Shott [-@shott_bipolar_1999] wedges are needed to work bone, antler, and wood and stone wedges are preferable than other types of material. For hunter-gatherers, these artifacts would provide multifunctionality from a single piece that could be continuously used until losing their morphological characteristics. And as previously mentioned, the fact that the presence of these artifacts is constant throughout Vale Boi’s Upper Paleolithic corroborates the importance that these pieces seem to have had. In addition, scaled pieces in Vale Boi may also be related to bone tool production, since Vale Boi is the Portuguese Upper Paleolithic site with more evidence for the production and use of bone tool technology [@evora_osseous_2013].

@@

Bipolar technology is quite common in Upper Paleolithic sites of Western Iberia, but current interpretations are based on empirical observations, rather than on more thoroughly analytical evidence. We argue that all factors mentioned in this study must be considered while interpreting each site in this region. While in Vale Boi there is clear evidence that the use of these artifacts would not be for bipolar knapping, in other sites the picture may be different.

>>>>

Bipolar technology clearly had an important role on the adaptive systems of the first modern humans in Western Iberia, and elsewhere [@villa_neandertals_2018]. A rise in bipolar technologies cannot be dissociated from the diverse set of factors that made Anatomically Modern Humans thrive. Changes in mobility patterns [@shott_diversity_2015], or the development of a "generalist specialist" ecological approach [@roberts_defining_2018], with particular emphasis on the diversification and intensification of resources use, are among some of those traits. It is in the contexts of this latter point that bipolar technology may have played a major role. In Vale Boi, since the earliest occupations at c. 32 ka cal BP, scaled pieces show up in the archaeological record connected to evidence related to an intensification and diversification of resource exploration, particularly the above-mentioned grease/marrow obtention techniques. These have been shown by the constant presence of impact fractures in ungulate bones, low percentage of bone areas related to higher amounts of grease, and the presence of red deer bone fragments from bones with considerably higher marrow and grease contents. Other indicators are both the constant presence of thermally altered quartz fragments that might be linked to stone boiling activities, and a large number of greywacke slabs with impact marks revealing their use as anvils. For carcass processing and bone marrow extraction the use of scaled pieces would allow for a better control of bone fracture. In fact, this would be much beteer then using a hard hammer directly on the bone, since this may either over fracture the bone or even crush it due to the lack of precision. The use of a wedge for these activities provides more control, avoiding complete crushing of the bone, making it much easier to cleanly extract bone marrow.

According to both Leblanc [-@leblanc_wedges_1992] and Shott [-@shott_bipolar_1999] wedges are needed to work bone, antler, and wood and stone wedges are preferable than other types of material. For hunter-gatherers, these artifacts would provide multifunctionality from a single piece that could be continuously used until losing their morphological characteristics. And as previously mentioned, the fact that the presence of these artifacts is constant throughout Vale Boi’s Upper Paleolithic corroborates the importance that these pieces seem to have had. In addition, scaled pieces in Vale Boi may also be related to bone tool production, since Vale Boi is the Portuguese Upper Paleolithic site with more evidence for the production and use of bone tool technology [@evora_osseous_2013].

Importantly, our results show that scaled pieces use at Vale Boi reveal techno-functional patterns that stay consistent across all paleolithic horizons.

One relevant trend is the fact that while quartz pieces were only explored in a single axis, chert pieces were, sometimes, explored in multiple axes. One possible, rather improbable, reason for these patterns might be the fact that quartz would be more easily available than chert. Another, more likely, possibility is due to the material properties of each raw material. While with quartz it may be possible to continuously use a single axis in a piece, reducing it in only one axis without losing its efficiency, with chert it may be necessary to rotate it in order to continue obtaining usable edges. On the other hand, the higher number of axis could simply suggest that chert was differently managed and economized. Nevertheless, the simple fact that most of the pieces with multiple axes are on average larger than the rest suggests that this hypothesis is likely incorrect. From an economic standpoint, the ideal would be to continuously use the same piece without any risk of loss of the features that made it ideal for this kind of use. Throughout the use of these pieces, they suffer successive violent blows, even the platform that rests on a surface suffers some sort of heavy impact. In this specific case, we argue that there is little to no control on how the piece’s features change over its lifetime use.

@@

Bipolar technology is quite common in Upper Paleolithic contexts of Western Iberia, but current interpretations are based on empirical observations, rather than on more solid analytical evidence. We argue that all factors mentioned in this study must be considered while interpreting each site in this region. While in Vale Boi there is clear evidence that the use of these artifacts would not be for bipolar knapping, in other sites the picture may be different.
